---
title: "Getting Started with glyph"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with glyph}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r setup}
library(glyph)
```

This vignette walks through glyph's core grammar with live, interactive
output. Every plot below is a real `glyph_spec` built with the package and
rendered to an actual D3-backed htmlwidget — not a screenshot.

One thing worth knowing up front: printing a `glyph_spec` at the console
auto-renders it (like a ggplot2 plot), but that auto-render only fires in an
interactive R session. Inside a vignette or pkgdown article the code runs
non-interactively, so each example below ends the pipeline with an explicit
[`render()`](../reference/render.html) call to produce the widget.

## 1. A basic scatterplot

No `aes()`, no `factor()` coercion — aesthetic mappings are just bare column
names passed straight to [`glyph()`](../reference/glyph.html) and
[`mark_point()`](../reference/mark_point.html).

```{r}
glyph(mtcars, x = wt, y = mpg) |>
  mark_point(color = cyl) |>
  render()
```

## 2. Tooltips and hover, declared in the pipeline

Interactivity is grammar, not glue. [`interact()`](../reference/interact.html)
turns on tooltips and a hover effect right where the plot is built, and
[`titles()`](../reference/titles.html) adds a title in the same pipe — no
`ggplotly()` conversion step, no lost formatting.

```{r}
glyph(mtcars, x = wt, y = mpg) |>
  mark_point(color = cyl) |>
  interact(tooltip = TRUE, hover = "enlarge") |>
  titles(title = "Motor Trend Cars") |>
  render()
```

Hover over a point to see it enlarge; hover longer to see the tooltip.

## 3. Animated bar chart

[`animate()`](../reference/animate.html) declares a transition as part of
the spec. `stagger` offsets each bar's entrance animation so they draw in
sequence rather than all at once.

```{r}
glyph(mtcars, x = cyl, y = mpg) |>
  mark_bar() |>
  animate(transition = "slide", stagger = 50) |>
  render()
```

Reload this page (or re-run the chunk in an R session) to see the bars
slide in.

## 4. Token-based dark theme

Instead of ggplot2's dozens of individual `theme()` arguments,
[`theme_tokens()`](../reference/theme_tokens.html) takes a small preset (or
individual tokens like `bg`, `font`, `accent`) and cascades foreground, grid,
and title colors automatically for contrast.

```{r}
glyph(mtcars, x = wt, y = mpg) |>
  mark_point(color = cyl) |>
  interact(tooltip = TRUE) |>
  theme_tokens(preset = "dark") |>
  titles(title = "Dark Theme Example") |>
  render()
```

## 5. Composed multi-plot layout

[`compose()`](../reference/compose.html) arranges multiple `glyph_spec`
objects into a single layout — here, two scatterplots side by side — without
reaching for `patchwork` or `cowplot`. Composed layouts are rendered the
same way as a single spec: pass the layout to `render()`.

```{r}
p1 <- glyph(mtcars, x = wt, y = mpg) |> mark_point(color = cyl)
p2 <- glyph(mtcars, x = hp, y = mpg) |> mark_point(color = cyl)

compose(p1, p2, type = "hstack") |>
  render()
```

## Where to next

- Browse the [function reference](../reference/index.html) for every mark,
  scale, and layout primitive glyph provides.
- Read [glyph vs ggplot2: Side-by-Side Comparison](comparison.html) for a
  broader tour of how the two grammars differ.
